\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-03-06 jasa/paper-jasa.tex
Accent

\if1\blind
{
  \title{\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers}
  \author{Kyle Cranmer$^1$, Juan Pavez$^2$ and Gilles Louppe$^1$\hspace{.2cm}\\
          $^1$New York University\\
          $^2$Federico Santa María University}
  \maketitle
} \fi

@@

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{1 - \hat
s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate estimates
of the ratio, even when using universal approximators such as a neural network
or a random forest. This observation is at the origin of this work, and led us
to show that the ratio $r(\mathbf{x})$ can still be computed exactly by forming
the equivalent statistic
$\frac{p(s(\mathbf{x})|\theta_0)}{p(s(\mathbf{x})|\theta_1)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\frac{\hat p(\hat s(\mathbf{x})|\theta_0)}{\hat p(\hat
s(\mathbf{x})|\theta_1)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be more accurate for all three classifiers, in particular in regions with enough density
and where the capacity of the
classifier is focused. By contrast, in regions where the densities are low
(i.e., at the boundaries) the approximated ratio show to less acurate
since the underlying classifier cannot make precise predictions, because
of the few training samples it is trained on in these parts of the input space.
When leveraging the fact that densities are mixtures and decomposing their
ratio, the capacity of the underlying classifiers can be focused on easier
supervised learning tasks, resulting as expected in even more accurate approximations, as
shown in cyan.

@@

    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat \gamma$.}
    \label{fig:2}
\end{figure}

Let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0$ versus the alternative $H_1: \gamma > 0$. That is, we want to evaluate
\begin{equation}
\Lambda({\cal D}; \Theta_0 = \{0\}, \Theta=(0, \frac{1}{2}]) =  \frac{ p({\cal D} | \gamma = 0) }{ \sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)} \; .
\end{equation}
Following the approach outlined in
Section~\ref{sec:generalized-likelihood-ratio}, computing the generalized
likelihood ratio statistic requires finding the maximum likelihood estimator
(MLE) for the denominator, which can be carried out using approximated
likelihood ratios. Reusing the decomposed neural network-based approximation and
minimizing the mean negative log-ratio of some observed data generated for
$\gamma=0.05$, Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs
over resampled observed data,
compared to the distribution of the true MLEs. The figure clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating estimator bias.  Figure \ref{fig:2b} also confirms
that the generalized likelihood ratio statistic obtained with our method is $\chi^2$ distributed,
as should be the true  generalized likelihood ratio statistic.

>>>>

\if1\blind
{
  \title{\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers}
  \author{Kyle Cranmer$^1$, Juan Pavez$^2$ and Gilles Louppe$^1$\hspace{.2cm}\\
          $^1$New York University\\
          $^2$Federico Santa Mar\'ia University}
  \maketitle
} \fi

@@

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat
s(\mathbf{x})} \right)$, as shown in green, may not always yield very accurate
estimates of the ratio, even when using universal approximators such as a neural
network or a random forest. This observation is at the origin of this work, and
led us to show that the ratio $r(\mathbf{x})$ can still be computed exactly by
forming the equivalent statistic
$\frac{p(s(\mathbf{x})|\theta_0)}{p(s(\mathbf{x})|\theta_1)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\frac{\hat p(\hat s(\mathbf{x})|\theta_0)}{\hat p(\hat
s(\mathbf{x})|\theta_1)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be more accurate for all three classifiers, even
when the direct approximation was not very good in the first place, e.g. as for the
linear model. In regions where the densities are low (i.e., at the
boundaries) the approximated ratio show to be less acurate since the underlying
classifier cannot make precise predictions, because of the few training samples
it is trained on in these parts of the input space. When leveraging the fact
that densities are mixtures and decomposing their ratio, the capacity of the
underlying classifiers can be focused on easier supervised learning tasks,
resulting as expected in even more accurate approximations, as shown in cyan.

@@

    \caption{Using approximated likelihood ratios for parameter inference yields an unbiased maximum likelihood estimator $\hat \gamma$, as empirically estimated 1000 times over resampled observed data.}
    \label{fig:2}
\end{figure}

Let us now consider the composite hypothesis testing setting, by comparing
$H_0: \gamma = 0$ versus the alternative $H_1: \gamma > 0$. That is, we want to evaluate
\begin{equation}
\Lambda({\cal D}; \Theta_0 = \{0\}, \Theta=(0, \frac{1}{2}]) =  \frac{ p({\cal D} | \gamma = 0) }{ \sup_{\theta \in \Theta} p({\cal D} | \gamma=\theta)} \; .
\end{equation}
Following the approach outlined in
Section~\ref{sec:generalized-likelihood-ratio}, computing the generalized
likelihood ratio statistic requires finding the maximum likelihood estimator
(MLE) for the denominator, which can be carried out using approximated
likelihood ratios. Reusing the decomposed neural network-based approximation and
minimizing the mean negative log-ratio of some observed data generated for
$\gamma=0.05$, Figure~\ref{fig:2a} shows the empirical distribution of the resulting MLEs,
compared to the distribution of the true MLEs. The figure clearly suggests that
the proposed inference procedure yields in this case an unbiased estimator
($\mathbb{E} [\hat \gamma] = 0.0501$, as empirically estimated over $1000$
trials). Note that had the underlying ratio not been accurate enough because of
poor training or poor calibration, the two distributions would not be exactly
overlapping, indicating estimator bias. As a second check, Figure \ref{fig:2b} also confirms
that the generalized likelihood ratio statistic obtained with our method is $\chi^2$ distributed,
as should be the true statistic.
